---
title: "Data analyses for Experiment 3"
author: "Scott Burgess"
date: "2024-09-15"
output: pdf_document
---

This code analyzes the data associated with Experiment 3 and produces Figure 4, 5, and 6 in the manuscript:

Barnes DK, Burgess SC. Fitness consequences of marine larval dispersal: the role of neighborhood density, spatial arrangement, and genetic relatedness on survival, growth, reproduction, and paternity. *Journal of Evolutionary Biology*

Code finalized September 2024.

Any comments or error reporting, please contact Scott Burgess: sburgess@bio.fsu.edu

## Load required libraries
```{r, message=F}
library('tidyverse')
library('glmmTMB')
library('gridExtra')
library('emmeans')
library('DHARMa')
library('vegan') 
```


```{r}
sessionInfo()
```


## Import data from Experiment 3

Phenotypic data

```{r}
dat <- read.csv("Data/Experiment_3.csv")
```

Paternity data

```{r}
Paternity_dat_0_9 <- read.table("Data/D0_9_final_Paternity.txt",header=T,sep=",")
```

Inferred fathers data

```{r}
BestCluster_0_9 <- read.csv("Data/D0_9_final_BestCluster.csv",header=T) 
```

Full sib family data 

```{r}
FSFamily_0_9 <- read.csv("Data/D0_9_final_BestFSFamily.csv")
```


# Process phenotypic data

Only use the summed data over all time periods

```{r}
dat <- dat %>% filter(Time_days=="all") %>% select(-Time_days, -Bifurcations, -Zooids)
```

Add ID for focal colonies
```{r}
focals <- c("A","B","D","G")
dat$focal <- ifelse(dat$Position %in% focals, 1, 0)
```

Create a vector of treatment levels and set the order (for plotting)

```{r}
treatment.vec <- c("alone","far","near","both")
dat$Treatment <- factor(dat$Treatment, levels=treatment.vec)
```


# Process paternity data

```{r}
# Add sample data to each data frame
brks <- seq(0,160,0.1)

add_sample_data <- function(d){
  d$MotherID <- rapply(strsplit(d$OffspringID,"_"),function(x) head(x,1))
  d$FatherID <- rapply(strsplit(d$InferredDad1,"_"),function(x) head(x,1))
  tmp <- dat[match(d$MotherID,dat$Colony),-1]
  names(tmp) <- paste0("Mother.",names(tmp))
  d <- cbind.data.frame(d,tmp)
  tmp <- dat[match(d$FatherID,dat$Colony),-1]
  names(tmp) <- paste0("Father.",names(tmp))
  d <- cbind.data.frame(d,tmp)
  
# Calculate euclidean distance between observed parents
  d$Distance <- NA
  for(i in 1:nrow(d)){
    d$Distance[i] <- dist(
      matrix(d[i,which(names(d) %in% c("Mother.X","Father.X","Mother.Y","Father.Y"))],2,2,byrow=T),method="euclidean")
  }
  
  return(d)
}

Paternity_dat_0_9 <- add_sample_data(Paternity_dat_0_9)
```

# Calculate the number of offspring sampled per mother

```{r}
OffspringIDs <- BestCluster_0_9$OffspringID
tmp <- rapply(strsplit(OffspringIDs,"_"),function(x) head(x,1))

MotherID <- paste0(tmp,"_parent")

offspring_per_mother <- as.data.frame(table(MotherID))
offspring_per_mother
```

# How many settlers were genotyped?
```{r}
with(offspring_per_mother,table(Freq))
nrow(offspring_per_mother) # mothers
sum(offspring_per_mother$Freq) # offspring
nrow(dat) # potential fathers
```

\

\


# Analyze Reproductive Output

```{r}
m1 <- glmmTMB(Offspring ~ Treatment + (1|Block), 
              ziformula = ~.,
              data = dat, 
              family = truncated_nbinom1)

m2 <- glmmTMB(Offspring ~ 1 + (1|Block), 
              ziformula = ~.,
              data = dat, 
              family = truncated_nbinom1)

result_reprod <- round(anova(m1, m2, test = "Chisq"), 3)
result_reprod
```

Check model fit

```{r, message=F}
simulateResiduals(fittedModel = m1, plot = T)
```

Get predicted values for plotting

```{r}
predictions_reprod = data.frame(Treatment = treatment.vec)

p <- emmeans(m1, ~ Treatment, 
             component = "cond", 
             type = "response")

predictions_reprod$fit <- summary(p)$response
predictions_reprod$lwr <- summary(p)$asymp.LCL
predictions_reprod$upr <- summary(p)$asymp.UCL
```

```{r}
nd <- expand.grid(Treatment = treatment.vec, 
                 Block = rownames(coef(m1)$cond$Block)) 

block_effects_reprod <- data.frame(nd,
                                fit = predict(m1,
                                              newdata = nd,
                                              type="response",
                                              se=F))
```

Reproductive output was similar for individuals in all treatments ($\chi^2$ = `r result_reprod$Chisq[2]`, df = `r result_reprod$'Chi Df'[2]`, p = `r result_reprod$'Pr(>Chisq)'[2]`).


\

\


# Analyze Relative Growth Rate

```{r}
m1 <- glmmTMB(rgr40 ~ Treatment + (1|Block), data = dat)

m2 <- glmmTMB(rgr40 ~ 1 + (1|Block), data = dat)

result_rgr <- round(anova(m1, m2, test = "Chisq"), 3)
result_rgr
```

Check model fit

```{r, message=F}
simulateResiduals(fittedModel = m1, plot = T)
```

Check influence of outliers

```{r}
outliers <- dat[which(resid(m1) < -0.06), 'Colony']
tmp <- dat %>% filter(!Colony %in% outliers)
m1 <- glmmTMB(rgr40 ~ Treatment + (1|Block), data = tmp)

m2 <- glmmTMB(rgr40 ~ 1 + (1|Block), data = tmp)

round(anova(m1, m2, test = "Chisq"), 3)
```

No qualitative change.

Relative growth rate was similar for individuals in all treatments ($\chi^2$ = `r result_rgr$Chisq[2]`, df = `r result_rgr$'Chi Df'[2]`, p = `r result_rgr$'Pr(>Chisq)'[2]`).

\

\


# Remove offspring from full sib families with Inclusion probabilities < 0.7

```{r}
tmp <- FSFamily_0_9 %>% filter(Prob.Inc.. < 0.7)

offspring_to_exclude <- c(tmp$Member1,tmp$Member2)

BestCluster_0_9 <- BestCluster_0_9 %>% filter(!(OffspringID %in% offspring_to_exclude))

```

### How many kept and excluded?
```{r}
nrow(FSFamily_0_9) # number of full sib families

nrow(FSFamily_0_9 %>% filter(Prob.Inc.. > 0.7)) # number of full sib families with high confidence

(nrow(FSFamily_0_9 %>% filter(Prob.Inc.. > 0.7)) / nrow(FSFamily_0_9)) * 100 # %

# number of full sib families with very high confidence (1)
nrow(FSFamily_0_9 %>% filter(Prob.Inc.. == 1)) 

(nrow(FSFamily_0_9 %>% filter(Prob.Inc.. == 1)) / nrow(FSFamily_0_9)) * 100 # %

nrow(tmp) # number full sib families excluded

length(offspring_to_exclude)  # number offspring excluded
```


```{r}
length(unique(BestCluster_0_9$MotherID)) # From this many mothers

nrow(BestCluster_0_9) # this many offspring

length(unique(BestCluster_0_9$FatherID)) # were sired by this many fathers

# this many of which were from outside the array
BestCluster_0_9 %>% filter(grepl("#",FatherID)) %>% summarize(n=n_distinct(FatherID)) 

# this many  offspring sired by colonies in the array
BestCluster_0_9 %>% filter(!(grepl("#",FatherID))) %>% summarize(n=n()) 

BestCluster_0_9 %>% group_by(FatherID) %>% 
  summarize(n=n_distinct(MotherID)) %>% count(n)
```


## Calculate the number of unique sires per focal colony

```{r, message=F}
n_fathers <- function(d){
  d$Mother_ID <- rapply(strsplit(d$MotherID,"_"),function(x) head(x,1))
  
  foo1 <- d %>% 
    group_by(Mother_ID,FatherID) %>% 
    summarise(n=n()) %>% 
    mutate(freq=n/sum(n)) %>% 
    ungroup()
  
  foo2 <- foo1 %>% group_by(Mother_ID) %>% 
    summarise(nGenotyped=sum(n),
              UniqueNumberFather=n_distinct(FatherID))
  
  foo2$standardized.n.father <- with(foo1, tapply(n, Mother_ID, vegan::rarefy, sample=19))
  foo2$standardized.n.father <- round(foo2$standardized.n.father,3)
    # add sample info 
  foo2 <- cbind.data.frame(foo2,dat[match(foo2$Mother_ID,dat$Colony),c(1:7,12)])
  
  foo2
}

n.fathers_0_9 <- n_fathers(d=BestCluster_0_9)
sort(n.fathers_0_9$standardized.n.father)
sort(n.fathers_0_9$UniqueNumberFather)
```

Calculate proportion for a beta glmm

```{r}
n.fathers_0_9$standardized.prop.father <- n.fathers_0_9$standardized.n.father / 20

# make the 1's 0.9999 so can fit beta glmm
n.fathers_0_9$standardized.prop.father <- ifelse(n.fathers_0_9$standardized.prop.father == 1,
                                                 0.99999,
                                                 n.fathers_0_9$standardized.prop.father)
```

# Analyze Number of Unique Sires per Mother

```{r}
m1 <- glmmTMB(standardized.prop.father  ~ Treatment + 
                (1|Block),
              family = 'beta_family',
              data = n.fathers_0_9)

m2 <- glmmTMB(standardized.prop.father  ~ 1 +
                (1|Block), 
              family = 'beta_family',
              data = n.fathers_0_9)

result_sires <- round(anova(m1, m2, test = "Chisq"), 3)
result_sires
```

Get predicted values for plotting

```{r}
predictions_sires = data.frame(Treatment = treatment.vec)

p <- emmeans(m1, ~ Treatment, 
             component = "cond", 
             type = "response")

predictions_sires$fit <- summary(p)$response * 20
predictions_sires$lwr <- summary(p)$asymp.LCL * 20
predictions_sires$upr <- summary(p)$asymp.UCL * 20
```

```{r}
nd <- expand.grid(Treatment = treatment.vec, 
                 Block = rownames(coef(m1)$cond$Block)) 

block_effects_sires <- data.frame(nd,
                                fit = predict(m1,
                                              newdata = nd,
                                              type="response",
                                              se=F))

block_effects_sires$fit <- block_effects_sires$fit * 20
```


The number of standardized unique sires per colony did not differ among treatments ($\chi^2$ = `r result_sires$Chisq[2]`, df = `r result_sires$'Chi Df'[2]`, p = `r result_sires$'Pr(>Chisq)'[2]`).


# Calculate the proportion of offspring from an assigned father

Get the mother ID's with an assigned father

```{r, message=F}
offspring.proportion <- function(a,b){
# Get the maternal ID for offspring with a father in the array 
    d1 <- a %>% 
    distinct(MotherID,.keep_all=T)
# Add mother and father ID's to BestCluster data frame
  b$Mother_ID <- rapply(strsplit(b$MotherID,"_"),function(x) head(x,1))
  b$Father_ID <- rapply(strsplit(b$FatherID,"_"),function(x) head(x,1))
# Get the data from BestCluster for offspring with a father in the array
# and calculate the number and frequency of offspring 
# per mother-father combination
  d2 <- b %>% 
    filter(Mother_ID %in% d1$MotherID) %>% 
    group_by(Mother_ID,Father_ID) %>% 
    summarise(n=n()) %>% 
    mutate(freq=n/sum(n)) %>% 
    ungroup()
# Calculate the max frequency of fathers per mother
  d3 <- d2 %>% 
    group_by(Mother_ID) %>% 
    mutate(max.freq=max(freq)) %>% 
    ungroup()
# Collect and arrange data  
  d4 <- d3[grep("#",d3$Father_ID,invert=T),]
  d5 <- cbind.data.frame(d4, d1[match(d4$Mother_ID,d1$MotherID),])
  d6 <- d5 %>% select(Mother_ID, Father_ID,
                      n, freq, max.freq,
                      ProbDad1,
                      Mother.Block, Mother.Treatment, Mother.Position, Mother.Direction,
                      Father.Block, Father.Treatment, Father.Position, Father.Direction)
  
  d6
}

offspring.proportion_0_9 <- offspring.proportion(Paternity_dat_0_9, BestCluster_0_9)
offspring.proportion_0_9
# Out of 511 offspring with full sib inclusion probabilities > 0.7, 
# 13 (2.5%) offspring were assigned paternity from one of 
# five candidate fathers in the experimental array
sum(offspring.proportion_0_9$n)
length(unique(offspring.proportion_0_9$Father_ID))
offspring.proportion_0_9 %>% count(Father_ID)
Paternity_dat_0_9 %>% count(Distance)
```

\

\

\


# Make Figure 4

```{r, fig.width = 6, fig.height = 2}

# Panel A
panel_A <- ggplot(data = predictions_reprod,
                  aes(x = Treatment,
                      y = fit)) +
  # geom_jitter(data = block_effects_reprod,
  #            aes(x = Treatment,
  #                y = fit),
  #            width = 0.05,
  #            alpha = 0.4,
  #            color = "grey") +
  geom_linerange(data = predictions_reprod,
                 aes(ymin = lwr,
                     ymax = upr)) +
  geom_point() +
  labs(y = "Reproductive output\n(number of settlers)",
       title = "a)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Panel B
panel_B <- ggplot(data = predictions_sires,
                  aes(x = Treatment,
                      y = fit)) +
  # geom_jitter(data = block_effects_sires,
  #            aes(x = Treatment,
  #                y = fit),
  #            width = 0.05,
  #            alpha = 0.4,
  #            color = "grey") +
  geom_linerange(aes(ymin = lwr,
                     ymax = upr)) +
  geom_point() +
  ylim(0, 20) +
  labs(y = "Standardized number\nof sires per colony",
       title = "b)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

# Panel C
tmp <- offspring.proportion_0_9 %>% filter(Mother.Block==Father.Block)

# far
y1 <- tmp %>% filter(Mother.Position=="B" & Father.Position == "C" |
                         Mother.Position=="C" & Father.Position == "B")
# near 
y2 <- tmp %>% filter(Mother.Position=="G" & Father.Position %in% c("H"))

# both
y3 <- tmp %>% filter(Mother.Position=="D" & Father.Position %in% c("E","F"))
  
tmp <- rbind.data.frame(y1, y2, y3)

panel_C <- ggplot(data = tmp, 
                  aes(x = Father.Treatment,
                      y = freq)) +
  geom_point() +
# for Mother_ID 02a, there were two fathers from the same block
    geom_linerange(data = tmp %>% filter(Mother_ID == "02a"),
                 aes(ymin = freq,
                     ymax = max.freq)) +
  labs(x = "Treatment",
       y = "Proportion of offspring\nsired by nearest colony",
       title = "c)") +
  scale_x_discrete(label = c("far","near","both\n(far)")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

grid.arrange(panel_A, panel_B, panel_C,
             nrow = 1, 
             ncol = 3)
```

```{r, echo=F}
pdf("Figure 4.pdf", width = 6, height = 2)

grid.arrange(panel_A, panel_B, panel_C,
             nrow = 1,
             ncol = 3)

dev.off()
```


# Make Figure 5 (Inclusion and exclusion probabilities)

```{r, fig.width = 4, fig.height = 2}
# Set breaks for plotting
brks <- seq(0,1,0.05)

# Get frequency of Inclusion probabilities
Inc <- with(FSFamily_0_9, hist(Prob.Inc.., 
                               breaks = brks, 
                               plot = F))

# Get frequency of Exclusion probabilities
Exc <- with(FSFamily_0_9,hist(Prob.Exc.., 
                              breaks = brks, 
                              plot = F))

# Panel A
d <- cbind.data.frame(mids = Inc$mids, counts = Inc$counts)
y <- d %>% filter(counts>0)

panel_A <- ggplot() +
  geom_col(data = y %>% filter(mids < 0.7), 
           aes(x = mids, 
               y = counts),
           fill = "grey",
           show.legend = F) + 
  xlim(0,1) + ylim(0,max(y$counts)) +
  geom_col(data = y %>% filter(mids >= 0.7), 
           aes(x = mids, 
               y = counts),
           fill = "black",
           show.legend = F) +
  labs(x = "Inclusion\nprobability",
       y = "Frequency of\nfull sib families",
       title = "a)") +
  theme_classic()
  
# Panel B
d <- cbind.data.frame(mids = Exc$mids, counts = Exc$counts)
y <- d %>% filter(counts>0)

panel_B <- ggplot() +
  geom_col(data = y, 
           aes(x = mids, 
               y = counts),
           fill = "black",
           show.legend = F) + 
  xlim(0,1) +
  labs(x = "Exlusion\nprobability",
       y = "",
       title = "b") +
  theme_classic()

grid.arrange(panel_A, panel_B, nrow = 1, ncol = 2)
```

```{r, echo=F}
pdf("Figure 5.pdf", width = 4, height = 2)

grid.arrange(panel_A, panel_B,
             nrow = 1,
             ncol = 2)

dev.off()
```

\

\


# Calculate the frequency of potential and observed distances between parents

```{r}
brks <- seq(0,200,0.1)
# # Potential distances
tmp <- dat %>% select("X", "Y")
mat <- dist(tmp, diag = T)
possible_distances <- mat[lower.tri(mat, diag = T)]
a <- hist(possible_distances, breaks = brks, plot = F)
possible_density <- data.frame(density = a$density / max(a$density),
                               mids = a$mids)
# Observed
density_freq <- function(d){
  b <- hist(d$Distance,breaks=brks,plot=F)
  data.frame(n = b$counts,
             density = b$density / max(b$density),
             mids = b$mids)
}

observed_density_0_9 <- density_freq(Paternity_dat_0_9)

possible_density$Metric <- "Potential"
observed_density_0_9$Metric <- "Observed"
density_data <- rbind.data.frame(possible_density, observed_density_0_9[,-1])
density_data$Metric <- factor(density_data$Metric, 
                              levels = c("Potential","Observed"))
```



# Make Figure 6

```{r, warning=F, fig.width = 4, fig.height = 3}

panel_A <- ggplot() +
  geom_col(data = density_data, 
                  aes(x = mids,
                      y = density,
                      fill = Metric),
           width = 3,
           alpha = 0.4,
           position_dodge(0.01)) +
  labs(x = "Distance between mother and father (m)", 
       y = "Relative\nfrequency",
       title = "a) Whole array") +
  scale_fill_manual(name = 'd',
                    breaks = c('Potential','Observed'),
                    values = c('Potential' = 'grey',
                               'Observed' = 'tomato')) +
  scale_x_continuous(breaks = seq(0,300,25)) +
  theme_classic() +
  theme(legend.position = 'none')

panel_B <- ggplot() +
  geom_col(data = density_data %>% filter(mids < 1.2), 
                  aes(x = mids,
                      y = density,
                      fill = Metric),
           width = 0.02,
           alpha = 0.4,
           position_dodge(0.1)) +
  labs(x = "Distance between mother and father (m)", 
       y = "Relative\nfrequency",
       title = "b) Within treatments") +
  scale_fill_manual(name = '',
                    breaks = c('Potential','Observed'),
                    values = c('Potential' = 'grey',
                               'Observed' = 'tomato')) +
  scale_x_continuous(breaks = seq(0,3,0.1)) +
  theme_classic() +
  theme(legend.position = c(0.8, 0.9),
        legend.key.size = unit(0.3, 'cm'))

grid.arrange(panel_A, panel_B,
             ncol = 1,
             nrow = 2)
```

```{r, echo=F, warning=F}
pdf("Figure 6.pdf", width = 4, height = 3)

grid.arrange(panel_A, panel_B,
             nrow = 2,
             ncol = 1)

dev.off()
```

