---
title: "Data analyses for Experiment 2"
author: "Scott Burgess"
date: "2024-09-15"
output: pdf_document
---

This code analyzes the data associated with Experiment 2 and produces Figure 3 in the manuscript:

Barnes DK, Burgess SC. Fitness consequences of marine larval dispersal: the role of neighborhood density, spatial arrangement, and genetic relatedness on survival, growth, reproduction, and paternity. *Journal of Evolutionary Biology*

Code finalized September 2024.

Any comments or error reporting, please contact Scott Burgess: sburgess@bio.fsu.edu

## Load required libraries
```{r, message=F}
library('tidyverse')
library('glmmTMB')
library('gridExtra')
library('emmeans')
```


```{r}
sessionInfo()
```


## Import data for Experiment 2
```{r}
dat <- read.csv("Data/Experiment_2.csv")
```

Create a new column called "Plate ID" for random effects
```{r}
dat$Plate.ID <- sub("_.*","",dat$Unique.ID) 
```

Create a vector of the ages at which data were recorded. Note that individuals were reared in the lab for 26 days, so age 46 days is 20 days in the field.
```{r}
Age.days.vec <- unique(dat$Age.days)
Age.days.vec
```

Select the data from when the colonies were retrieved from the field after 20 days (age 46 days)
```{r}
d <- dat %>% filter(Age.days == Age.days.vec[2]) 
```

# Survival

Generalized linear mixed effects model with a binomial distribution and random effect of plate number (Plate.ID).

```{r}
survival1 <- glmmTMB(Survival ~ Relatedness +
                         (1|Plate.ID), 
                     family = "binomial",
                     data = d)

survival2 <- glmmTMB(Survival ~ 1 +
                         (1|Plate.ID),
                     family = "binomial",
                     data = d)
```


Test Relatedness
```{r}
result1 <- round(anova(survival1, survival2,test="Chisq"), 3)
result1
```

There was no evidence that relatedness affected survival ($\chi^2$ = `r result1$Chisq[2]`, df = `r result1$'Chi Df'[2]`, p = `r result1$'Pr(>Chisq)'[2]`).


### Calculate fitted values for plotting

```{r}
predictions_survival <- data.frame(Relatedness = unique(dat$Relatedness))

p <- predict(survival1, 
             predictions_survival, 
             type = "link",
             se.fit = T,
             re.form = NA)

predictions_survival$fit <- plogis(p$fit)
predictions_survival$se.fit <- p$se.fit
# Calculate confidence intervals
predictions_survival$lwr <- plogis(p$fit - 2 * p$se.fit)
predictions_survival$upr <- plogis(p$fit + 2 * p$se.fit)
```



\

\


# Relative growth rate

Change data to the wide format

```{r}
dat_wide <- dat %>% pivot_wider(id_cols=Unique.ID:Position,
                                names_from=Age.days, 
                                values_from=c(Survival,Bifurcations,Zooids))
```

Calculate relative growth rate 

```{r}
dat_wide$relgrowth <- (log(dat_wide$Zooids_46)- log(dat_wide$Zooids_26)) / (46-26)
```


Create a new column called "Plate ID" for random effects

```{r}
dat_wide$Plate.ID <- sub("_.*","",dat_wide$Unique.ID) 
```

Fit generalized linear mixed effects model with a Gaussian distribution and random effect of plate number (Unique.ID). 

```{r}
rgr1 <- glmmTMB(relgrowth ~ Relatedness +
                         (1|Plate.ID), 
                     family = "gaussian",
                     data = dat_wide)

rgr2 <- glmmTMB(relgrowth ~ 1 +
                         (1|Plate.ID),
                     family = "gaussian",
                     data = dat_wide)
```

Test Relatedness

```{r}
result1 <- round(anova(rgr1, rgr2,test="Chisq"), 3)
result1
```

There was no evidence that relatedness affected relative growth rate ($\chi^2$ = `r result1$Chisq[2]`, df = `r result1$'Chi Df'[2]`, p = `r result1$'Pr(>Chisq)'[2]`).


### Calculate fitted values for plotting

```{r}
predictions_rgr <- data.frame(Relatedness = unique(dat$Relatedness))

p <- predict(rgr1, 
             predictions_rgr, 
             type = "link",
             se.fit = T,
             re.form = NA)

predictions_rgr$fit <- p$fit
predictions_rgr$se.fit <- p$se.fit
# Calculate confidence intervals
predictions_rgr$lwr <- p$fit - 2 * p$se.fit
predictions_rgr$upr <- p$fit + 2 * p$se.fit
```

``` {r}
nd <- data.frame(Relatedness = c(rep("Non",5),rep("Sib",10)), 
                 Plate.ID = rownames(coef(rgr1)$cond$Plate.ID)) 

plate_effects_rgr <- data.frame(nd,
                                fit = predict(rgr1,
                                              newdata = nd,
                                              type="response",
                                              se=F))
```


# Reproductive output

Add zero ovicells for colonies that died
```{r}
d$Ovicells.total <- ifelse(is.na(d$Ovicells.total),0,d$Ovicells.total)
```


Fit zero-inflated generalized linear mixed effects model with a truncated negative binomial distribution and random effect of plate number (Unique.ID). 

```{r}
totov1 <- glmmTMB(Ovicells.total ~ Relatedness + 
                      (1|Plate.ID), 
                  ziformula = ~.,
                  data = d, 
                  family = truncated_nbinom1)

totov2 <- glmmTMB(Ovicells.total ~ 1 +
                      (1|Plate.ID), 
                  ziformula = ~.,
                  data = d, 
                  family = truncated_nbinom1)
```


```{r}
result1 <- round(anova(totov1, totov2,test="Chisq"), 3)
result1
```

There was no evidence that relatedness affected reproductive output ($\chi^2$ = `r result1$Chisq[2]`, df = `r result1$'Chi Df'[2]`, p = `r result1$'Pr(>Chisq)'[2]`).

### Calculate fitted values for plotting

```{r}
predictions_totov <- data.frame(Relatedness = unique(dat$Relatedness))

# p <- predict(totov1, 
#              predictions_totov, 
#              type = "link",
#              se.fit = T,
#              re.form = NA)
# 
# predictions_totov$fit <- exp(p$fit)
# predictions_totov$se.fit <- p$se.fit
# # Calculate confidence intervals
# predictions_totov$lwr <- exp(p$fit - 2 * p$se.fit)
# predictions_totov$upr <- exp(p$fit + 2 * p$se.fit)

p <- emmeans(totov1, ~ Relatedness, 
             component = "cond", 
             type = "response")

predictions_totov$fit <- summary(p)$response
predictions_totov$lwr <- summary(p)$asymp.LCL
predictions_totov$upr <- summary(p)$asymp.UCL

```

```{r}
nd <- data.frame(Relatedness = c(rep("Non",5),rep("Sib",10)), 
                 Plate.ID = rownames(coef(totov1)$cond$Plate.ID)) 

plate_effects_totov <- data.frame(nd,
                                fit = predict(totov1,
                                              newdata = nd,
                                              type="response",
                                              se=F))
plate_effects_totov <- d %>% 
    group_by(Plate.ID) %>% 
    summarize(fit = mean(Ovicells.total))

plate_effects_totov$Relatedness <- ifelse(
    grepl("non", plate_effects_totov$Plate.ID),
    "Non",
    "Sib")
```


# Make Figure 3

```{r, fig.width=6, fig.height=2}
panel_A <- ggplot(data = predictions_survival, 
                  aes(x = Relatedness,
                      y = fit)) +
    geom_linerange(aes(min = lwr,
                       max = upr)) +
    geom_point() +
    ylim(0,1) +
    labs(y = "Probability of survival",
         title = "a)") +
    scale_x_discrete(labels = c("Unrelated","Siblings")) +
    theme_classic()

panel_B <- ggplot(data = predictions_rgr, 
                  aes(x = Relatedness,
                      y = fit)) +
    # geom_jitter(data = plate_effects_rgr, 
    #            aes(x = Relatedness,
    #                y = fit),
    #            alpha = 0.4,
    #            width = 0.05) +
    geom_linerange(aes(min = lwr,
                       max = upr)) +
    geom_point() +
    ylim(0,0.3) +
    labs(y = "Relative growth rate",
         title = "b)") +
    scale_x_discrete(labels = c("Unrelated","Siblings")) +
    theme_classic()

panel_C <- ggplot(data = predictions_totov, 
                  aes(x = Relatedness,
                      y = fit)) +
    geom_linerange(aes(min = lwr,
                       max = upr)) +
    geom_point() +
    ylim(0,1600) +
    labs(y = "Reproductive output",
         title = "c)") +
    scale_x_discrete(labels = c("Unrelated","Siblings")) +
    theme_classic()

grid.arrange(panel_A, panel_B, panel_C, 
             nrow = 1,
             ncol = 3)
```

```{r, echo=F}
pdf("Figure 3.pdf", width = 6, height = 2)

grid.arrange(panel_A, panel_B, panel_C, 
             nrow = 1,
             ncol = 3)

dev.off()
```



