---
title: "Data analyses for Experiment 1"
author: "Scott Burgess"
date: "2024-09-15"
output: pdf_document
---

This code analyzes the data associated with Experiment 1 and produces Figure 2 in the manuscript:

Barnes DK, Burgess SC. Fitness consequences of marine larval dispersal: the role of neighborhood density, spatial arrangement, and genetic relatedness on survival, growth, reproduction, and paternity. *Journal of Evolutionary Biology*

Code finalized September 2024.

Any comments or error reporting, please contact Scott Burgess: sburgess@bio.fsu.edu

## Load required libraries
```{r, message=F}
library('tidyverse')
library('glmmTMB')
library('gridExtra')
```


```{r}
sessionInfo()
```


## Import data for Experiment 1
```{r}
dat <- read.csv("Data/Experiment_1.csv")
```

Create a vector of the ages at which data were recorded. Note that individuals were reared in the lab for 17 days, so age 25 days is 8 days in the field, and age 32 days is 15 days in the field
```{r}
Age.days.vec <- unique(dat$Age.days)
Age.days.vec
```


# Survival

Fit generalized linear mixed effects model with a binomial distribution and random effect of plate number (Unique.ID). 

## Survival after 8 days in the field

```{r}
d <- dat %>% filter(Age.days == Age.days.vec[2])

survival1 <- glmmTMB(Survival ~ Density * Relatedness + 
                         (1|Unique.ID), 
                     family = "binomial", 
                     data = d) 

survival2 <- glmmTMB(Survival ~ Density + Relatedness + 
                         (1|Unique.ID), 
                     family = "binomial",
                     data = d)

survival3 <- glmmTMB(Survival ~ Density +
                         (1|Unique.ID), 
                     family = "binomial", 
                     data = d)

survival4 <- glmmTMB(Survival ~ Relatedness +
                         (1|Unique.ID), 
                     family = "binomial", 
                     data = d)
```

Test the interaction
```{r}
result1 <- round(anova(survival1, survival2,test="Chisq"), 3)
result1
```

Test `relatedness`
```{r}
result2 <- round(anova(survival2, survival3,test="Chisq"), 3)
result2
```

Test `density`
```{r}
result3 <- round(anova(survival2, survival4,test="Chisq"), 3)
result3
odds.effect <- round((1 - exp(confint(survival2)[2,])) * 100, 2)
```

Results text:

The probability of individual colony survival in the field declined with density after 8 days ($\chi^2$ = `r result3$Chisq[2]`, df = `r result3$'Chi Df'[2]`, p = `r result3$'Pr(>Chisq)'[2]`).

The odds of survival declined by `r odds.effect[3]` (`r odds.effect[2]` - `r odds.effect[1]`) for every additional neighbor colony.

There was no evidence that relatedness had additive ($\chi^2$ = `r result2$Chisq[2]`, df = `r result2$'Chi Df'[2]`, p = `r result2$'Pr(>Chisq)'[2]`) or interactive effects ($\chi^2$ = `r result1$Chisq[2]`, df = `r result1$'Chi Df'[2]`, p = `r result1$'Pr(>Chisq)'[2]`) on survival. 


### Calculate fitted values for plotting

```{r}
predictions_survival_8days <- expand.grid(
    Density = seq(0, 20, 0.1),
    Relatedness = unique(dat$Relatedness))

p <- predict(survival2, 
             predictions_survival_8days, 
             type = "link",
             se.fit = T,
             re.form = NA)

predictions_survival_8days$fit <- plogis(p$fit)
predictions_survival_8days$se.fit <- p$se.fit
# Calculate confidence intervals
predictions_survival_8days$lwr <- plogis(p$fit - 2 * p$se.fit)
predictions_survival_8days$upr <- plogis(p$fit + 2 * p$se.fit)
```



## Survival after 15 days in the field

```{r}
d <- dat %>% filter(Age.days == Age.days.vec[3])

survival1 <- glmmTMB(Survival ~ Density * Relatedness + 
                         (1|Unique.ID), 
                     family = "binomial", 
                     data = d) 

survival2 <- glmmTMB(Survival ~ Density + Relatedness + 
                         (1|Unique.ID), 
                     family = "binomial",
                     data = d)

survival3 <- glmmTMB(Survival ~ Density +
                         (1|Unique.ID), 
                     family = "binomial", 
                     data = d)

survival4 <- glmmTMB(Survival ~ Relatedness +
                         (1|Unique.ID), 
                     family = "binomial", 
                     data = d)
```

Test the interaction
```{r}
result1 <- round(anova(survival1, survival2,test="Chisq"), 3)
result1
```

Test `relatedness`
```{r}
result2 <- round(anova(survival2, survival3,test="Chisq"), 3)
result2
```

Test `density`
```{r}
result3 <- round(anova(survival2, survival4,test="Chisq"), 3)
result3
odds.effect <- round((1 - exp(confint(survival2)[2,])) * 100, 2)
```

Results text:

The probability of individual colony survival in the field declined with density after 15 days ($\chi^2$ = `r result3$Chisq[2]`, df = `r result3$'Chi Df'[2]`, p = `r result3$'Pr(>Chisq)'[2]`).

The odds of survival declined by `r odds.effect[3]` (`r odds.effect[2]` - `r odds.effect[1]`) for every additional neighbor colony.

There was no evidence that relatedness had additive ($\chi^2$ = `r result2$Chisq[2]`, df = `r result2$'Chi Df'[2]`, p = `r result2$'Pr(>Chisq)'[2]`) or interactive effects ($\chi^2$ = `r result1$Chisq[2]`, df = `r result1$'Chi Df'[2]`, p = `r result1$'Pr(>Chisq)'[2]`) on survival. 

### Calculate fitted values for plotting

```{r}
predictions_survival_15days <- expand.grid(
    Density = seq(0, 20, 0.1),
    Relatedness = unique(dat$Relatedness))

p <- predict(survival2, 
             predictions_survival_15days, 
             type = "link",
             se.fit = T,
             re.form = NA)

predictions_survival_15days$fit <- plogis(p$fit)
predictions_survival_15days$se.fit <- p$se.fit
# Calculate confidence intervals
predictions_survival_15days$lwr <- plogis(p$fit - 2 * p$se.fit)
predictions_survival_15days$upr <- plogis(p$fit + 2 * p$se.fit)
```


\

\

\


# Relative growth rate

Change data to the wide format

```{r}
dat_wide <- dat %>% pivot_wider(id_cols = Unique.ID:Inside.outside,
                                names_from = Age.days, 
                                values_from = c(Survival, Bifurcations,Zooids))
```

Calculate relative growth rate 

8 days

```{r}
dat_wide$relgrowth.collection1 <- (log(dat_wide$Zooids_25)- log(dat_wide$Zooids_17)) / (25-17)
```

15 days

```{r}
dat_wide$relgrowth.collection2 <- (log(dat_wide$Zooids_32)- log(dat_wide$Zooids_17)) / (32-17)
```

Fit generalized linear mixed effects model with a Gaussian distribution and random effect of plate number (Unique.ID). 

## Relative growth rate after 8 days in the field

```{r}
rgr1 <- glmmTMB(relgrowth.collection1 ~ Density * Relatedness +
                    (1|Unique.ID), 
                data = dat_wide, 
                family = gaussian())

rgr2 <- glmmTMB(relgrowth.collection1 ~ Density + Relatedness + 
                    (1|Unique.ID), 
                data = dat_wide,
                family = gaussian())

rgr3 <- glmmTMB(relgrowth.collection1 ~ Density +
                    (1|Unique.ID), 
                data = dat_wide,
                family = gaussian())

rgr4 <- glmmTMB(relgrowth.collection1 ~ Relatedness +
                    (1|Unique.ID), 
                data = dat_wide,
                family = gaussian())
```

Test the interaction
```{r}
result1 <- round(anova(rgr1, rgr2,test="Chisq"), 3)
result1
```

Test `relatedness`
```{r}
result2 <- round(anova(rgr2, rgr3,test="Chisq"), 3)
result2
```

Test `density`
```{r}
result3 <- round(anova(rgr2, rgr4,test="Chisq"), 3)
result3
```

Results text:

No evidence that density ($\chi^2$ = `r result3$Chisq[2]`, df = `r result3$'Chi Df'[2]`, p = `r result3$'Pr(>Chisq)'[2]`) or relatedness ($\chi^2$ = `r result2$Chisq[2]`, df = `r result2$'Chi Df'[2]`, p = `r result2$'Pr(>Chisq)'[2]`) had additive or interactive effects ($\chi^2$ = `r result1$Chisq[2]`, df = `r result1$'Chi Df'[2]`, p = `r result1$'Pr(>Chisq)'[2]`) on relative growth rate. 


### Calculate fitted values for plotting

```{r}
predictions_rgr_8days <- expand.grid(
    Density = seq(0, 20, 0.1),
    Relatedness = unique(dat$Relatedness))

p <- predict(rgr2, 
             predictions_rgr_8days, 
             type = "link",
             se.fit = T,
             re.form = NA)

predictions_rgr_8days$fit <- p$fit
predictions_rgr_8days$se.fit <- p$se.fit
# Calculate confidence intervals
predictions_rgr_8days$lwr <- p$fit - 2 * p$se.fit
predictions_rgr_8days$upr <- p$fit + 2 * p$se.fit
```


## Relative growth rate after 15 days in the field

```{r}
rgr1 <- glmmTMB(relgrowth.collection2 ~ Density * Relatedness +
                    (1|Unique.ID), 
                data = dat_wide, 
                family = gaussian())

rgr2 <- glmmTMB(relgrowth.collection2 ~ Density + Relatedness + 
                    (1|Unique.ID), 
                data = dat_wide,
                family = gaussian())

rgr3 <- glmmTMB(relgrowth.collection2 ~ Density +
                    (1|Unique.ID), 
                data = dat_wide,
                family = gaussian())

rgr4 <- glmmTMB(relgrowth.collection2 ~ Relatedness +
                    (1|Unique.ID), 
                data = dat_wide,
                family = gaussian())
```

Test the interaction
```{r}
result1 <- round(anova(rgr1, rgr2,test="Chisq"), 3)
result1
```

Test `relatedness`
```{r}
result2 <- round(anova(rgr2, rgr3,test="Chisq"), 3)
result2
```

Test `density`
```{r}
result3 <- round(anova(rgr2, rgr4,test="Chisq"), 3)
result3
```

Results text:
No evidence that density ($\chi^2$ = `r result3$Chisq[2]`, df = `r result3$'Chi Df'[2]`, p = `r result3$'Pr(>Chisq)'[2]`) or relatedness ($\chi^2$ = `r result2$Chisq[2]`, df = `r result2$'Chi Df'[2]`, p = `r result2$'Pr(>Chisq)'[2]`) had additive or interactive effects ($\chi^2$ = `r result1$Chisq[2]`, df = `r result1$'Chi Df'[2]`, p = `r result1$'Pr(>Chisq)'[2]`) on relative growth rate. 


### Calculate fitted values for plotting

```{r}
predictions_rgr_15days <- expand.grid(
    Density = seq(0, 20, 0.1),
    Relatedness = unique(dat$Relatedness))

p <- predict(rgr2, 
             predictions_rgr_15days, 
             type = "link",
             se.fit = T,
             re.form = NA)

predictions_rgr_15days$fit <- p$fit
predictions_rgr_15days$se.fit <- p$se.fit
# Calculate confidence intervals
predictions_rgr_15days$lwr <- p$fit - 2 * p$se.fit
predictions_rgr_15days$upr <- p$fit + 2 * p$se.fit
```





# Make Figure 2
```{r, fig.width=5, fig.height=5}
my_colors <- c("#98c1d9","#ee6c4d")
legend.labs <- c("Unrelated","Siblings")

panel_A <- ggplot(data = predictions_survival_8days, 
                  aes(x = Density,
                      y = fit,
                      group = Relatedness)) + 
    geom_ribbon(aes(ymin = lwr,
                    ymax = upr,
                    fill = Relatedness),
                alpha = 0.4) +
    geom_line(aes(color = Relatedness)) +
    scale_color_manual(values = my_colors,
                       labels = legend.labs) +
    scale_fill_manual(values = my_colors,
                       labels = legend.labs) +
    ylim(0,1) +
    labs(x = expression(paste("Density (colonies per 12", cm^2, ")")),
         y = "Probability of survival",
         title = "a) 8 days in the field") +
    theme_classic() +
    theme(legend.position = c(0.3, 0.25),
          legend.title=element_blank(),
          legend.background = element_rect(fill='transparent'))

panel_B <- ggplot(data = predictions_survival_15days, 
                  aes(x = Density,
                      y = fit,
                      group = Relatedness)) + 
    geom_ribbon(aes(ymin = lwr,
                    ymax = upr,
                    fill = Relatedness),
                alpha = 0.4) +
    geom_line(aes(color = Relatedness)) +
    scale_color_manual(values = my_colors,
                       labels = legend.labs) +
    scale_fill_manual(values = my_colors,
                       labels = legend.labs) +
    ylim(0,1) +
    labs(x = expression(paste("Density (colonies per 12", cm^2, ")")),
         y = "Probability of survival",
         title = "b) 15 days in the field") +
    theme_classic() +
    theme(legend.position = c(0.3, 0.25),
          legend.title=element_blank(),
          legend.background = element_rect(fill='transparent'))

panel_C <- ggplot(data = predictions_rgr_8days, 
                  aes(x = Density,
                      y = fit,
                      group = Relatedness)) + 
    geom_ribbon(aes(ymin = lwr,
                    ymax = upr,
                    fill = Relatedness),
                alpha = 0.4) +
    geom_line(aes(color = Relatedness)) +
    scale_color_manual(values = my_colors,
                       labels = legend.labs) +
    scale_fill_manual(values = my_colors,
                       labels = legend.labs) +
    ylim(0,0.25) +
    labs(x = expression(paste("Density (colonies per 12", cm^2, ")")),
         y = "Relative growth rate",
         title = "c) 8 days in the field") +
    theme_classic() +
    theme(legend.position = c(0.3, 0.25),
          legend.title=element_blank(),
          legend.background = element_rect(fill='transparent'))


panel_D <- ggplot(data = predictions_rgr_15days, 
                  aes(x = Density,
                      y = fit,
                      group = Relatedness)) + 
    geom_ribbon(aes(ymin = lwr,
                    ymax = upr,
                    fill = Relatedness),
                alpha = 0.4) +
    geom_line(aes(color = Relatedness)) +
    scale_color_manual(values = my_colors,
                       labels = legend.labs) +
    scale_fill_manual(values = my_colors,
                       labels = legend.labs) +
    ylim(0,0.25) +
    labs(x = expression(paste("Density (colonies per 12", cm^2, ")")),
         y = "Relative growth rate",
         title = "d) 15 days in the field") +
    theme_classic() +
    theme(legend.position = c(0.3, 0.25),
          legend.title=element_blank(),
          legend.background = element_rect(fill='transparent'))

grid.arrange(panel_A, panel_B, panel_C, panel_D, 
             nrow = 2,
             ncol = 2)
```

```{r, echo=F}
pdf("Figure 2.pdf", width = 5, height = 5)

grid.arrange(panel_A, panel_B, panel_C, panel_D, 
             nrow = 2,
             ncol = 2)

dev.off()
```


